---
title: "STAT420 Homework 4"
author: "Alok K. Shukla"
date: "9/18/2016"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: readable
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Solutions

## Exercise 1 (Using `lm` for Inference)

For this exercise we will again use the `faithful` dataset. Remember, this is a default dataset in `R`, so there is no need to load it. You should use `?faithful` to refresh your memory about the background of this dataset about the duration and waiting times of eruptions of [the Old Faithful geyser](http://www.yellowstonepark.com/about-old-faithful/) in [Yellowstone National Park](https://en.wikipedia.org/wiki/Yellowstone_National_Park).

**(a)** Fit the following simple linear regression model in `R`. Use the eruption duration as the response and waiting time as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `faithful_model`. Use a $t$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.01$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**Solution**

```{r}
faithful_model = lm(eruptions~waiting,data=faithful)
faithful_model_test_info = summary(faithful_model)$coefficients
beta_0_hat      = faithful_model_test_info[1,1] # Estimate
beta_0_hat_t    = faithful_model_test_info[1,3] # t value
beta_0_hat_pval = faithful_model_test_info[1,4] # Pr(>|t|)

beta_1_hat      = faithful_model_test_info[2,1] # Estimate
beta_1_hat_t    = faithful_model_test_info[2,3] # t value
beta_1_hat_pval = faithful_model_test_info[2,4] # Pr(>|t|)
```
*The null and alternative hypotheses*

For $\beta_0$
 
$H_0:\beta_0=0$
$H_1:\beta_0!=0$
 
 
For $\beta_1$
 
$H_0:\beta_1=0$
$H_1:\beta_1!=0$


*The value of the test statistic*

For $\beta_0$

```{r}
beta_0_hat_t
```

 
For $\beta_1$
```{r}
beta_1_hat_t
```
 

*The p-value of the test*

For $\beta_0$
```{r}
beta_0_hat_pval
```
 
For $\beta_1$
```{r}
beta_1_hat_pval
```

 
*A statistical decision at $\alpha = 0.01$*

For $\beta_0$
```{r}
alpha = 0.01
if(beta_0_hat_pval > alpha){
  print("Fail to reject NULL hypothesis")
}else{
  print("Reject NULL hypothesis")
}
```

For $\beta_1$
```{r}
alpha = 0.01
if(beta_1_hat_pval> alpha){
  print("Fail to reject NULL hypothesis")
}else{
  print("Reject NULL hypothesis")
}
```


*A conclusion in the context of the problem*

Since we succcessfully reject the NULL hypothesis for both $\beta_0$ and $\beta_1$ ,we say there is a significant linear relationship between eruptions duration and waiting time.




**(b)** Calculate a 99% confidence interval for $\beta_1$. Give an interpretation of the interval in the context of the problem.

**Solution**

```{r}
confint(faithful_model, parm = "waiting", level = 0.99)
```

We are 99% confident that for an increase in waiting time of 1 minute, the average increase in eruption duration is between 0.0698727 and 0.0813832 minutes.


**(c)** Calculate a 90% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.


**Solution**

```{r}
confint(faithful_model, parm = "(Intercept)", level = 0.90)
```

We are 90% confident that the average duration of an eruption waiting 0 minutes is between -2.138335 and -1.609697 minutes, but we don’t really believe that, since a zero waiting time is essentially saying, the eruption hasnt really ended.



**(d)** Use a 95% confidence interval to estimate the mean eruption duration for waiting times of 75 and 80 minutes. Which of the two intervals is wider? Why?

**Solution**

```{r}
new_waiting = data.frame(waiting = c(75, 80))
res = predict(faithful_model, newdata = new_waiting, 
        interval = c("confidence"), level = 0.95)

res

if ((res[1,"upr"] - res[1,"lwr"]) > (res[2,"upr"] - res[2,"lwr"])){
  print("Interval for 75 minutes is wider.")
}else{
  print("Interval for 85 minutes is wider.")
  }

```

The confidence interval for 85 minutes is larger because we are less certain about our predictions as the waiting time increases.

**(e)** Use a 95% prediction interval to predict the eruption duration for waiting times of 75 and 100 minutes.

**Solution**

```{r}
new_waiting = data.frame(waiting = c(75, 100))
predict(faithful_model, newdata = new_waiting, 
        interval = c("prediction"), level = 0.95)
```


**(f)** Create a scatterplot of the data. Add the regression line, 95% confidence bands, and 95% prediction bands.

**Solution**

```{r}
waiting_grid = seq(min(faithful$waiting), max(faithful$waiting), by = 0.01)
eruptions_ci_band = predict(faithful_model, 
                           newdata = data.frame(waiting = waiting_grid), 
                           interval = "confidence", level = 0.95)
eruptions_pi_band = predict(faithful_model, 
                           newdata = data.frame(waiting = waiting_grid), 
                           interval = "prediction", level = 0.95) 
plot(eruptions~waiting,data=faithful,
     xlab = "Waiting time (in Minutes)",
     ylab = "Eruption Duration (in Minutes)",
     main = "Waiting time vs Eruption Duration",
     pch  = 20,
     cex  = 2,
     col  = "dodgerblue")

abline(faithful_model, lwd = 5, col = "darkorange")

lines(waiting_grid, eruptions_ci_band[,"lwr"], col = "red", lwd = 3, lty = 2)
lines(waiting_grid, eruptions_ci_band[,"upr"], col = "red", lwd = 3, lty = 2)
lines(waiting_grid, eruptions_pi_band[,"lwr"], col = "green", lwd = 3, lty = 3)
lines(waiting_grid, eruptions_pi_band[,"upr"], col = "green", lwd = 3, lty = 3)
legend("topleft", c("Confidence","Prediction"),lty=c(1,1),lwd=c(2.5,2.5),col=c("red","green")) 
```


## Exercise 2 (Using `lm` for Inference)

For this exercise we will again use the `diabetes` dataset, which can be found in the `faraway` package.

**(a)** Fit the following simple linear regression model in `R`. Use the total cholesterol as the response and weight as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `cholesterol_model`. Use an $F$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The ANOVA table (You may use `anova()` and omit the row for Total.)
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**Solution**

```{r}
library(faraway)
cholesterol_model = lm(chol~weight,data=diabetes)
```

*The null and alternative hypotheses*


 
$H_0:\beta_1=0$

Total cholesterol does not depend on weight.
 
$H_1:\beta_1!=0$
 
Total cholesterol does depend on weight.
 

*The ANOVA table*




```{r}
anova_res = anova(cholesterol_model)
anova_res["weight",]
```

*The value of the test statistic*

```{r}
anova_res["weight",4]
``` 

*The p-value of the test*


```{r}
p_val = anova_res["weight",5]
p_val
```
 

 
*A statistical decision at $\alpha = 0.05$*


```{r}
alpha = 0.05
if(p_val > alpha){
  print("Fail to reject NULL hypothesis")
}else{
  print("Reject NULL hypothesis")
}
```



*A conclusion in the context of the problem*

Since we failed to reject the NULL hypothesis ,we can not say there is a significant linear relationship between weight and cholesterol levels.


**(b)** Fit the following simple linear regression model in `R`. Use HDL as the response and weight as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `hdl_model`. Use an $F$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
- The ANOVA table (You may use `anova()` and omit the row for Total.)
- The value of the test statistic
- The p-value of the test
- A statistical decision at $\alpha = 0.05$
- A conclusion in the context of the problem

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**Solution**

```{r}
hdl_model = lm(hdl~weight,data=diabetes)
```

*The null and alternative hypotheses*


 
$H_0:\beta_1=0$

Total hdl does not depend on weight.
 
$H_1:\beta_1!=0$
 
Total hdl does depend on weight.
 

*The ANOVA table*




```{r}
anova_res = anova(hdl_model)
anova_res["weight",]
```

*The value of the test statistic*

```{r}
anova_res["weight",4]
``` 

*The p-value of the test*


```{r}
p_val = anova_res["weight",5]
p_val
```
 

 
*A statistical decision at $\alpha = 0.05$*


```{r}
alpha = 0.05
if(p_val > alpha){
  print("Fail to reject NULL hypothesis")
}else{
  print("Reject NULL hypothesis")
}
```



*A conclusion in the context of the problem*

Since we successfully rejected the NULL hypothesis ,we can say there is a significant linear relationship between hdl levels and weight.



## Exercise 3 (Inference "without" `lm`)

For this exercise we will once again use the data stored in [`goalies.csv`](goalies.csv). It contains career data for all 716 players in the history of the National Hockey League to play goaltender through the 2014-2015 season. The two variables we are interested in are:

- `W` - Wins
- `MIN` - Minutes

Fit a SLR model with `W` as the response and `MIN` as the predictor. Test $H_0: \beta_1 = 0.008$ vs $H_1: \beta_1 < 0.008$ at $\alpha = 0.01$. Report the following: 

- $\hat{\beta_1}$
- $SE[\hat{\beta_1}]$
- The value of the $t$ test statistic
- The degrees of freedom
- The p-value of the test
- A statistical decision at $\alpha = 0.01$

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

You should use `lm()` to fit the model and obtain the estimate and standard error. But then you should directly calculate the remaining values. Hint: be careful with the degrees of freedom. Think about how many observations are being used.

**Solution**

```{r}
goalies <- read.csv("goalies.csv")
goal_model = lm(W~MIN,data = goalies)
goal_model_test_info = summary(goal_model)$coefficients
beta_1_hat      = goal_model_test_info[2,1] # Estimate
beta_1_hat_se   = goal_model_test_info[2,2] # Std. Error
```
*$\hat{\beta_1}$*

```{r}
beta_1_hat
```

*$SE[\hat{\beta_1}]$*

```{r}
beta_1_hat_se
```

*The value of the $t$ test statistic*

```{r}
calculated_t = (beta_1_hat - 0.008) / beta_1_hat_se
calculated_t
```

*The degrees of freedom*

```{r}
df = length(resid(goal_model)) - 2
df
```

*The p-value of the test*

```{r}
calculated_p = 2 * pt(abs(calculated_t), df = df, lower.tail = TRUE)
calculated_p
```

*A statistical decision at $\alpha = 0.01$*

```{r}
alpha = 0.01
if(calculated_p > alpha){
  print("Fail to reject NULL hypothesis")
}else{
  print("Reject NULL hypothesis")
}
```


## Exercise 4 (Simulating Sampling Distributions)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 4$
- $\beta_1 = 0.5$
- $\sigma^2 = 25$

We will use samples of size $n = 50$.

**(a)** Simulate this model $1500$ times. Each time use `lm()` to fit a SLR model, then store the value of $\hat{\beta}_0$ and $\hat{\beta}_1$. Set a seed using **your** UIN before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
uin = 658325691
set.seed(uin)
n = 50
x = seq(0, 20, length = n)
```

**Solution**

```{r}
Sxx = sum((x - mean(x)) ^ 2)
beta_0 = 4
beta_1 = 0.5
sigma  = sqrt(25)
(var_beta_1_hat = sigma ^ 2 / Sxx)
(var_beta_0_hat = sigma ^ 2 * (1 / n + mean(x) ^ 2 / Sxx))

num_samples = 1500
beta_0_hats = rep(0, num_samples)
beta_1_hats = rep(0, num_samples)

for(i in 1:num_samples){
  eps = rnorm(n, mean = 0, sd = sigma)
  y   = beta_0 + beta_1 * x + eps
  
  sim_model = lm(y ~ x)
  
  beta_0_hats[i] = coef(sim_model)[1]
  beta_1_hats[i] = coef(sim_model)[2]
}
```


**(b)** For the *known* values of $x$, what is the expected value of $\hat{\beta}_1$?

**Solution**

```{r}
beta_1
```


**(c)** For the known values of $x$, what is the standard deviation of $\hat{\beta}_1$?

**Solution**

```{r}
( sd_beta_1_hat = sqrt(var_beta_1_hat) ) 
```


**(d)** What is the mean of your simulated values of $\hat{\beta}_1$? Does this make sense given your answer in **(b)**?

**Solution**

```{r}
mean(beta_1_hats)
```

Its almost equal to the the expected value of $\hat{\beta}_1$ for known x; which makes sense.

**(e)** What is the standard deviation of your simulated values of $\hat{\beta}_1$? Does this make sense given your answer in **(c)**?

**Solution**

```{r}
sd(beta_1_hats)
```
Its almost equal to the standard deviation of $\hat{\beta}_1$ for known x, which makes sense.

**(f)** For the known values of $x$, what is the expected value of $\hat{\beta}_0$?

**Solution**

```{r}
beta_0
```


**(g)** For the known values of $x$, what is the standard deviation of $\hat{\beta}_0$?

**Solution**

```{r}
( sd_beta_0_hat = sqrt(var_beta_0_hat) ) 
```


**(h)** What is the mean of your simulated values of $\hat{\beta}_0$? Does this make sense given your answer in **(f)**?

**Solution**

```{r}
mean(beta_0_hats)
```

Its almost equal to the the expected value of $\hat{\beta}_0$ for known x; which makes sense.

**(i)** What is the standard deviation of your simulated values of $\hat{\beta}_0$? Does this make sense given your answer in **(g)**?

**Solution**

```{r}
sd(beta_0_hats)
```
Its almost equal to the standard deviation of $\hat{\beta}_0$ for known x, which makes sense.


**(j)** Plot a histogram of your simulated values for $\hat{\beta}_1$. Add the normal curve for the true sampling distribution of $\hat{\beta}_1$.

**Solution**

```{r}
hist(beta_1_hats, prob = TRUE, breaks = 20, 
     xlab = expression(hat(beta)[1]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_1, sd = sqrt(var_beta_1_hat)), 
      col = "darkorange", add = TRUE, lwd = 3)
```


**(k)** Plot a histogram of your simulated values for $\hat{\beta}_0$. Add the normal curve for the true sampling distribution of $\hat{\beta}_0$.

**Solution**

```{r}
hist(beta_0_hats, prob = TRUE, breaks = 20, 
     xlab = expression(hat(beta)[0]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_0, sd = sqrt(var_beta_0_hat)),
      col = "darkorange", add = TRUE, lwd = 3)
```

## Exercise 5 (Simulating Confidence Intervals)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 1$
- $\beta_1 = 3$
- $\sigma^2 = 16$

We will use samples of size $n = 20$.

Our goal here is to use simulation to verify that the confidence intervals really do have their stated confidence level.

**(a)** Simulate this model $2000$ times. Each time use `lm()` to fit a SLR model, then store the value of $\hat{\beta}_0$ and $s_e$. Set a seed using **your** UIN before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
uin = 658325691
set.seed(uin)
n = 20
x = seq(-5, 5, length = n)
```
**Solution**

```{r}
Sxx = sum((x - mean(x)) ^ 2)
beta_0 = 1
beta_1 = 3
sigma  = sqrt(16)
(var_beta_1_hat = sigma ^ 2 / Sxx)
(var_beta_0_hat = sigma ^ 2 * (1 / n + mean(x) ^ 2 / Sxx))

num_samples = 2000
beta_0_hats = rep(0, num_samples)
se_hats = rep(0, num_samples)
for(i in 1:num_samples){
  eps = rnorm(n, mean = 0, sd = sigma)
  y   = beta_0 + beta_1 * x + eps
  
  sim_model = lm(y ~ x)
  
  beta_0_hats[i] = coef(sim_model)[1]
  se_hats[i] = summary(sim_model)$sigma 
  
}
```


**(b)** For each of the $\hat{\beta}_0$ that you simulated calculate a 90% confidence interval. Store the lower limits in a vector `lower_90` and the upper limits in a vector `upper_90`. Some hints:

- You will need to use `qt()` to calculate the critical value, which will be the same for each interval.
- Remember that `x` is fixed, so $S_{xx}$ will be the same for each interval.
- You could, but do not need to write a `for` loop. Remember vectorized operations.

**Solution**

```{r}
lower_90 = rep(0, num_samples)
upper_90 = rep(0, num_samples)

temp = 1-(0.1/2)
critical_val <- qt(temp,df=n-2)
for ( i in 1:num_samples){
  lower_90[i] = beta_0_hats[i] - 
              critical_val*se_hats[i]*sqrt((1/n)+((mean(x)^2)/Sxx))
  upper_90[i] = beta_0_hats[i] + critical_val*se_hats[i]*sqrt((1/n)+((mean(x)^2)/Sxx))
 
}
  

```

**(c)** What proportion of these intervals contain the true value of $\beta_0$?

**Solution**

```{r}
count = 0;
for (i in 1:num_samples)
  if (beta_0<=upper_90[i] & beta_0>=lower_90[i]){
    count = count+1;
  }
count
(count/num_samples)
```


**(d)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_0 = 0$ vs $H_1: \beta_0 \neq 0$ at $\alpha = 0.10$?

**Solution**

```{r}
count = 0;
beta_00 = 0
for (i in 1:num_samples){
  if (beta_00>upper_90[i] | beta_00<lower_90[i]){
    count = count+1;
  }
}
(count/num_samples)

```


**(e)** For each of the $\hat{\beta}_0$ that you simulated calculate a 99% confidence interval. Store the lower limits in a vector `lower_99` and the upper limits in a vector `upper_99`.

**Solution**
```{r}
lower_99 = rep(0, num_samples)
upper_99 = rep(0, num_samples)
temp = 1-(0.01/2)
critical_val <- qt(temp,df=n-2)
for ( i in 1:num_samples){
  lower_99[i] = beta_0_hats[i] - 
              critical_val*se_hats[i]*sqrt((1/n)+((mean(x)^2)/Sxx))
  upper_99[i] = beta_0_hats[i] + critical_val*se_hats[i]*sqrt((1/n)+((mean(x)^2)/Sxx))
}
 
```




**(f)** What proportion of these intervals contain the true value of $\beta_0$?

**Solution**

```{r}
count = 0;
for (i in 1:num_samples)
  if (beta_0<=upper_99[i] & beta_0>=lower_99[i]){
    count = count+1;
  }
count
(count/num_samples)
```


**(g)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_0 = 0$ vs $H_1: \beta_0 \neq 0$ at $\alpha = 0.01$?

**Solution**

```{r}
count = 0;
beta_00 = 0
for (i in 1:num_samples)
  if (beta_00>upper_99[i] | beta_00<lower_99[i]){
    count = count+1;
  }

(count/num_samples)
```


