---
title: "STAT420 Homework 1"
author: "Alok K. Shukla"
date: "8/28/2016"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment Solutions

### Exercise 1 (Working with Vectors)
The given vector of data.

```{r}
x = c(1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, 144)
```

**(a)** Calculation of the sample mean of `x` *without* the use of `mean()`, `sd()`, `sum()`, or `+`.

**Solution**
Using the matrix multiplication operator `%*%`

If we multiply a row vector with a unit column vector of conformable size, we will essentially get sum of elements in row vector.

1. Create a row matrix of `x` using Transpose function `t` on Matrix function `matrix`
```{r}
x_rowmatrix = t(matrix(x))
x_rowmatrix
```
2. Create a unit vector of conformable size
```{r}
unit_vector = rep(1, times = length(x))
unit_vector
```
3. Multiply both and get the value of total using `as.numeric`
```{r}
total = as.numeric(x_rowmatrix %*% unit_vector)
total
```
Note: This works even when the second argument is a vector. Help doc for `%*%` says *"If one argument is a vector, it will be promoted to either a row or column matrix to make the two arguments conformable"*

4. Get the mean
```{r}
computed_mean = total/length(x)
computed_mean
```

5. The mean computed with `mean`

```{r}
mean(x)
```
```{r}

```

**(b)** Calculate the sample standard deviation of `x` *without* the use of `mean()`, `sd()`, `sum()`, or `+`.

**Solution**

If we take the mod of cross product of a vector with itself we essentially get the sum of squares of the elements in the vector. By definition SD is sqaure root of average sum of residual ( actual value $-$ mean).

1. Create the residual vector
```{r}
x_residuals = x - computed_mean
x_residuals
```

2. Compute sum of squares
```{r}
ssq_x = as.numeric(crossprod(x_residuals))
ssq_x
```

3. Compute average and square root
```{r}
computed_sd = sqrt(ssq_x/(length(x)-1))
computed_sd
```

4. Compare with actual sd
```{r}
sd(x)
```

### Exercise 2 (Vectors and Matrices)

**(a)** Create five vectors `x0`, `x1`, `x2`, `x3`, and `y`. Each should have a length of 30 and store the following:

- `x0`: Each element should be the value `1`.

**Solution**
```{r}
x0 = rep(1,30)
x0
```

- `x1`: The first 30 square numbers.

**Solution**
```{r}
x1 = (seq(1,30,1))^2
x1
length(x1)
```

- `x2`: 30 evenly spaced numbers between `0` and `1`.

**Solution**
```{r}
x2 = (seq(0,1,(1/29)))
x2
length(x2)
```

- `x3`: The natural log of the integers from `1` to `30`.

**Solution**
```{r}

x3 = log(seq(1,30,1))
x3
```

- `y`: The result of running the following code, after creating the other four vectors:

```{r, eval = FALSE}
set.seed(42)
y  = 5 * x0 + 1 * x1 + 6 * x2 + 3 * x3 + rnorm(n = 30, mean = 0 , sd = 1)
```
Report the value returned from `sum(y)`.

**Solution**
```{r, echo=FALSE}
set.seed(42)
y  = 5 * x0 + 1 * x1 + 6 * x2 + 3 * x3 + rnorm(n = 30, mean = 0 , sd = 1)
```
```{r}
sum(y)
```

**(b)** Create a matrix `X` which stores `x0`, `x1`, `x2`, and `x3` as columns of the matrix. Report the value returned from `sum(X)`.

**Solution**
```{r}
X = cbind(x0,x1,x2,x3)
sum(X)
```

**(c)** Use matrix operations to create a new matrix `beta_hat` defined as follows:

\[
\hat{\beta} = (X^T X)^{-1} X^T y.
\]

Report the values stored in the matrix `beta_hat`. Note that you can use the vector `y` as if it is a $30 \times 1$ matrix.

**Solution**
```{r}
beta_hat = (solve(t(X)%*%X))%*%(t(X)%*%y)
beta_hat
```

**(d)** Subset `beta_hat` matrix to exclude the first row, then square each element and report the sum of these results.

**Solution**
```{r}
beta_hat_subset_squaresum = as.numeric(crossprod(beta_hat[2:4,]))
beta_hat_subset_squaresum
```

**(e)** Create a new variable `y_hat` which stores the result of the matrix operation,

\[
\hat{y} = X \hat{\beta}.
\]

Perform and report the result of the following operation,

\[
\sum_{i = 1}^{30} (y_i - \hat{y}_i)^ 2.
\]

**Solution**
```{r}
y_hat = X%*%beta_hat
residual_y = y-y_hat
ssq_y = as.numeric(crossprod(residual_y))
ssq_y
```

### Exercise 3 (Discrete Probability)

The [2016 United States presidential election](https://en.wikipedia.org/wiki/United_States_presidential_election,_2016) has been an interesting contest. A [Fox News National Poll](http://www.foxnews.com/politics/interactive/2016/06/09/fox-news-poll-national-releas-6-16/) from June which surveyed 1004 registered voters placed former New Mexico Governor [Gary Johnson](https://en.wikipedia.org/wiki/Gary_Johnson) in third place, which is expected, however polling in the double digits at 12%. ([Third party candidates haven't performed well in recent years. The last time a third party managed to get votes in the Electoral College was 1968.](https://en.wikipedia.org/wiki/United_States_presidential_election,_1968))

Suppose the true proportion of registered voters that support Johnson is 12% and we obtain our own random sample of 50 registered voters. Answer the follows questions as they relate to this sample of 50 voters.

**(a)** What is the probability that exactly 5 of the voters are Johnson supporters?

**Solution**
```{r}
dbinom(5,size=50,prob=0.12)
```

**(b)** What is the probability that 10 or fewer voters are Johnson supporters?

**Solution**
```{r}
pbinom(10,size=50,prob=0.12,lower.tail = TRUE)
```

**(c)** What is the probability that 37 or more voters are **not** Johnson supporters?

**Solution**

```{r, message=FALSE}
require(Rmpfr)
one = mpfr(1,120)
prob = mpfr(pbinom(37,size=50,prob=0.12,lower.tail = FALSE),120)
one - prob
```

**(d)** What is the probability that between 3 and 9 (inclusive) voters are Johnson supporters?

**Solution**

```{r}
numbers = seq(3,9)
prob = dbinom(numbers,size=50,prob=0.12,log=FALSE)
sum(prob)
```

### Exercise 4 (Continuous Probability)

For this exercise, consider a random variable $X$ which is normally distributed with a mean of 120 and a standard deviation of 15. That is,

\[
X \sim N\left(\mu = 120, \sigma^2 = 225\right).
\]

**(a)** Calculate $P(X < 95)$.

**Solution**

```{r}
pnorm(95,mean=120,sd=15,lower.tail = TRUE,log.p=FALSE)-dnorm(95,mean=120,sd=15)
```

**(b)** Calculate $P(X > 140)$.

**Solution**

```{r}
pnorm(140,mean=120,sd=15,lower.tail = FALSE,log.p=FALSE)
```

**(c)** Calculate $P(95 < X < 120)$.

**Solution**

```{r}
pnorm(120,mean=120,sd=15,lower.tail = TRUE,log.p=FALSE) - pnorm(95,mean=120,sd=15,lower.tail = TRUE,log.p=FALSE)
```

**(d)** Find $q$ such that $P(X < q) = 0.05$.

**Solution**

```{r}
qnorm(0.05,mean=120,sd=15,lower.tail = TRUE,log.p=FALSE)
```

**(e)** Find $q$ such that $P(X > q) = 0.10$.

**Solution**

```{r}
qnorm(0.10,mean=120,sd=15,lower.tail = FALSE,log.p=FALSE)
```




