---
title: "STAT420 Homework 2"
author: "Alok K. Shukla"
date: "09/05/2016"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Solutions

## Exercise 1 (Writing Simple Functions)

Available vectors

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
a = 1:10
b = 10:1
c = rep(1, times = 10)
d = 2 ^ (1:10)
```

**(a)** Write a function called `sum_of_squares`.

- Arguments:
    - A vector of numeric data `x`.
- Output:
    - The sum of the squares of the elements of the vector. $\sum_{i = 1}^n x_i^2$
    
Provide the function, as well as the result of running the following code:

```{r, eval = FALSE}
sum_of_squares(x = a)
sum_of_squares(x = c(c, d))
```

**Solution**

We use the `crossprod` function as discussed in previous assignment.

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sum_of_squares = function(x){
  as.numeric(crossprod(x))
}
```

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sum_of_squares(x = a)

sum_of_squares(x = c(c, d))
```

**(b)** Write a function called `sum_of_power`.

- Arguments:
    - A vector of numeric data `x`.
    - `p` which should have the default value of `2`.
- Output:
    - $\sum_{i = 1}^n x_i^{p}$
    
Provide the function, as well as the result of running the following code:

```{r, eval = FALSE}
sum_of_power(x = a)
sum_of_power(x = a, p = 3)
sum_of_power(x = a, p = a)
sum_of_power(x = a, p = c(1, 2))
```

**Solution**

Using `^` operator.


```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sum_of_power = function(x,p=2){
  sum(x^p)
}
```

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

sum_of_power(x = a)

sum_of_power(x = a, p = 3)

sum_of_power(x = a, p = a)

sum_of_power(x = a, p = c(1, 2))
```

**(c)** Write a function called `rms_diff`.

- Arguments:
    - A vector of numeric data `x`.
    - A vector of numeric data `y`.
- Output:
    - $\sum_{i = 1}^n (x_i - y_i)^{2}$
    
Provide the function, as well as the result of running the following code:

```{r, eval = FALSE}
rms_diff(x = a, y = b)
rms_diff(x = d, y = c)
rms_diff(x = d, y = 1)
rms_diff(x = a, y = 0) ^ 2 * length(a)
```

**Solution**

Standard operations.


```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
rms_diff = function(x,y){
  sum((x-y)^2)
}
```

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
rms_diff(x = a, y = b)

rms_diff(x = d, y = c)

rms_diff(x = d, y = 1)

rms_diff(x = a, y = 0) ^ 2 * length(a)
```

## Exercise 2 (Plotting, Testing)

For this exercise we will use the data that is [stored in `intelligence.csv`](intelligence.csv) which records IQs of a random sample of residents of Pawnee and Eagleton, Indiana.

**(a)** Load the data from `intelligence.csv` into a variable in `R` called `intelligence`. Show the code used to do this. 

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
intelligence <- read.csv("intelligence.csv")
```


**(b)** Create a side-by-side boxplot that compares the IQs across the two towns. Be sure to give the plot a title and label the axes appropriately.

**Solution**

Using `ggplot2` library's methods. Setting Aesthetics and Title parameters.

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}

library(ggplot2)

ggplot(intelligence, aes(x=town, y=iq)) + geom_boxplot(fill='#A4A4A4', color="darkblue") + ggtitle("Comparing IQs of two towns") 


```

**(c)** Are people from Eagleton smarter than people from Pawnee? Perform an appropriate statistical test using the given sample data. That is, test $H_0: \mu_E = \mu_P$ vs $H_1: \mu_E > \mu_P$, where

- $\mu_E$ is the mean IQ of a resident of Eagleton.
- $\mu_P$ is the mean IQ of a resident of Pawnee.

Explicitly state the p-value of the test and the resulting statistical decision at a significance level $\alpha = 0.10$. Interpret the results in the context of the problem.

**Solution**
The null hypothesis

 $H_0: \mu_E = \mu_P$

Alternate hypothesis

 $H_1: \mu_E > \mu_P$
 
```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
pawnee = intelligence[intelligence$town=="pawnee",1]
eagleton = intelligence[intelligence$town=="eagleton",1]
res = t.test(eagleton, pawnee, alternative = c("greater"), var.equal = TRUE)
res$p.value
```
We now have the p-value of our test, which is greater than our significance level (0.10), so we fail to reject the null hypothesis implying we cant say if Eagletonians are smarter than Pawnee people.


**(d)** Repeat **(c)** using a two-sided alternative hypothesis. What changes?

**Solution**

 
```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

res2 = t.test(eagleton, pawnee, alternative = c("two.sided"), var.equal = TRUE)
res2$p.value
```

The p-value increases, we still fail to reject null hypothesis.

## Exercise 3 (Writing More Functions)

In this exercise we will write our own functions related to performing a one-sample $t$ test. That is $H_{0}: \mu = \mu_{0}$ versus $H_{1}: \mu \neq \mu_{0},$ where $\mu_{0}$ is the hypothesized value of $\mu$.


**(a)** Write a function called `do_t_test` which takes two inputs:

- `x`: A vector which stores observations.
- `mu`: The hypothesized value of $\mu$ which defaults to `0`.

The function should output:

- The value of the test statistic, $t$.
- The p-value of the test. The function only needs to be able to handle a two-sided alternative.


**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
do_t_test = function(x,mu=0){
  x_bar = mean(x)
  s = sd(x) 
  mu_0 = mu
  n= length(x)
  t = (x_bar - mu_0) / (s / sqrt(n))
  pval = 2*pt(-abs(t),df=n-1) ## two tailed
  c(t, pval)
}

```


**(b)** Write a function called `make_decision` which takes two inputs:

- `pval`: The p-value of a test.
- `alpha`: The significance level of a test. Set a default value of `0.05`.

The function should output `"Reject!"` or `"Fail to Reject."` based on the comparison of `pval` to `alpha`.

**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
make_decision = function(pval,alpha){
  ifelse(pval>alpha,"Fail to Reject.","Reject.")
}

```

**(c)** Now we will test the quality of your functions from parts **(a)** and **(b)**. Run the following code:

```{r, eval = FALSE}
set.seed(42)
y = rnorm(25, 1.4, 1)
pval = do_t_test(y, mu = 2)[2]
pval
make_decision(pval, alpha = 0.10)
```


**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
set.seed(42)
y = rnorm(25, 1.4, 1)
pval = do_t_test(y, mu = 2)[2]
pval
make_decision(pval, alpha = 0.10)
```





## Exercise 4 (CLT Simulation)

For this exercise we will simulate from the exponential distribution. If a random variable $X$ has an exponential distribution with rate parameter $\lambda$, the pdf of $X$ can be written

\[
f(x;\lambda) = \lambda e ^ {-\lambda x}
\]

for $x \geq 0$.


**(a)** This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. 

```{r}
birthday = 18760613
set.seed(birthday)
```

**Solution**

```{r}
birthday = 19920120
set.seed(birthday)
```


**(b)** Simulate 10000 samples of size **5** from an exponential distribution with $\lambda = 2$. Store the mean of each sample in a vector. Plot a histogram of these sample means. (Be sure to give it a title, and label the axes appropriately.) Based on the histogram, do you think the central limit theorem applies here?

**Solution**

We use `apply` method with a custom function to replace the traditional `for` loop that re-assigns values in a vector. 

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

x = data.frame(rep(0,10000))
x_bars = apply(x, 1, function(x) mean(rexp(5,2)))


x_bar_hist = hist(x_bars, breaks = 50, 
                  main = "Histogram of Sample Means",
                  xlab = "Sample Means")

```

We can see from the histogram that the distribution of sample means is not normal; its right tail is longer. Thus, the Central Limit Theorem doesn't hold.

**(c)** Simulate 10000 samples of size **100** from an exponential distribution with $\lambda = 2$. Store the mean of each sample in a vector. Plot a histogram of these sample means. (Be sure to give it a title, and label the axes appropriately.) Based on the histogram, do you think the central limit theorem applies here?

**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

x = data.frame(rep(0,10000))
x_bars = apply(x, 1, function(x) mean(rexp(100,2)))


x_bar_hist = hist(x_bars, breaks = 50, 
                  main = "Histogram of Sample Means",
                  xlab = "Sample Means")

```


We can see from the histogram that the distribution of sample means is normal; there is no skewness. Thus, the Central Limit Theorem holds.


**(d)** We just repeated ourselves, so that means we probably should be writing a function. Write a function called `sim_xbars_exp` which takes three inputs:

- The number of samples to simulate.
- The sample size.
- The rate parameter of an exponential distribution.

The function should output a vector of sample means which are the result of sampling from an exponential distribution as specified by the inputs.

Use your function to simulate 25000 samples of size **50** from an exponential distribution with $\lambda = 3$. Store the mean of each sample in a vector. Plot a histogram of these sample means.

**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sim_xbars_exp = function(no_of_iter, sample_size, rate_param){
  x = data.frame(rep(0,no_of_iter))
  apply(x, 1, function(x) mean(rexp(sample_size,rate_param)))
}


x_bar_hist = hist(sim_xbars_exp(25000,50,3), breaks = 50,
            main = "Histogram of Sample Means",
            xlab = "Sample Means")

```

## Exercise 5 (More Simulation)

Let $X$ follow an exponential distribution with rate parameter $\lambda_X = 2$. Let $Y$ follow a Poisson distribution with rate parameter $\lambda_Y = 3$.

We write $sd(X)$ for the true standard deviation of $X$ and $m(Y)$ for the true median of $Y$.

Let $s_x$ be the sample standard deviation of $X$ which is an estimate of $sd(X)$. Also let $m_y$ be the sample median which is an estimate of $m(Y)$.

Suppose we take samples of size $n_x = 10$ from $X$ and take samples of size $n_y = 5$. Consider the statistic

\[
\frac{s_x}{m_y}.
\]

Simulate $\frac{s_x}{m_y}$ 5000 times and store the results. Plot a histogram of the observed values of $\frac{s_x}{m_y}$. Comment on the shape of the histogram and empirical distribution of $\frac{s_x}{m_y}$. Before running your code, set the same seed used for the previous exercise. 

**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
set.seed(birthday)

N = data.frame(rep(0,5000))
s_x = apply(N, 1, function(N) sd(rexp(10,2)))
m_y = apply(N, 1, function(N) median(rpois(5,3)))
new_var = s_x/m_y
new_var_hist = hist(new_var, breaks = 50,
  main = "Histogram of Ratio of Sample SD and Sample Median",
  xlab = "Sample SD/Sample Median")


```

The histogram doesnt like that of a Normal Distribution. It looks more like a Gamma distribution. We can try to fit it in one as following.

```{r,eval=FALSE}
install.packages("fitdistrplus")
```

```{r,message=F, warning=F}
library("fitdistrplus")
```

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

## Remove Inf from z_bars
new_var_clean = new_var[new_var!=Inf]
fit.gamma = fitdist(new_var_clean, "gamma")
fit.gamma
plot(fit.gamma)
gofstat(fit.gamma)
```

Lets try to plot sample means of `new_var` and check if Central Limit Theorem holds.

First by re-generating using the fitted `gamma` distribution. A Gamma distribution is defined by `shape` and `rate`; which we populate from fitted `fit.gamma`.

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

x = data.frame(rep(0,10000))
x_bars = apply(x, 1, function(x) mean(rgamma(as.numeric(fit.gamma$estimate["shape"]),as.numeric(fit.gamma$estimate["rate"]))))


x_bar_hist = hist(x_bars, breaks = 50, 
main = "Histogram of Sample Means (Generated using Gamma dist.)",
xlab = "Sample Means")

```


 Now with resampling the orginal distribution of `new_var` using the `sample` method ( sample size =100 ).
 
```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}

x = data.frame(rep(0,10000))
x_bars = apply(x, 1, function(x) mean(sample(new_var, 100, replace = TRUE)))


x_bar_hist = hist(x_bars, breaks = 50, 
  main = "Histogram of Sample Means (Bootstrapped)",
  xlab = "Sample Means")

```

We thus conclude that the Central Limit Theorem holds.

