---
title: "STAT420 Homework 3"
author: "Alok K. Shukla"
date: "9/12/2016"
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment Solutions

## Exercise 1 (Using `lm`)

For this exercise we will use the `faithful` dataset. This is a default dataset in `R`, so there is no need to load it. You should use `?faithful` to learn about the background of this dataset.

**(a)** Suppose we would like to predict the duration of an eruption of [the Old Faithful geyser](http://www.yellowstonepark.com/about-old-faithful/) in [Yellowstone National Park](https://en.wikipedia.org/wiki/Yellowstone_National_Park) based on the waiting time before an eruption. Fit a simple linear model in `R` that accomplishes this task. Store the results in a variable called `faithful_model`. Output the result of calling `summary()` on `faithful_model`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
faithful_model = lm(eruptions~waiting, data = faithful)
summary(faithful_model)
```


**(b)** Output only the estimated regression coefficients. Interpret $\beta_0$ and $\hat{\beta_1}$ in the *context of the problem*. Be aware that only one of those is an estimate.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
faithful_model["coefficients"]
```
$\beta_0$ tells us the mean duration of an eruption that occurred without any waiting. The estimated value of this is negative, doesnt seem right though.

$\hat{\beta_1} = 0.0756$  tells us that for an increase in waiting time of one minute, the estimated mean duration of eruption increases by $0.075$ minutes.


**(c)** Use your model to predict the duration of an eruption based on a waiting time of **80** minutes. Do you feel confident in this prediction? Briefly explain.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
predict(faithful_model,data.frame(waiting = 80))
```

I feel confident about this prediction, looking at the actual values in `faithful`, the predicted value seems about right. Thats becuase the predictor is in range.

**(d)** Use your model to predict the duration of an eruption based on a waiting time of **120** minutes. Do you feel confident in this prediction? Briefly explain.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
predict(faithful_model,data.frame(waiting = 120))
```

I dont feel confident about this prediction, looking at the actual values in `faithful`, I can say nothing about its accuracy. Thats becuase the predictor is not in range.

**(e)** Calculate the RSS for this model.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sum(faithful_model$residuals^2)
sum((faithful_model$fitted.values - faithful$eruptions)^2)
```


**(f)** Create a scatterplot of the data and add the fitted regression line. Make sure your plot is well labeled and is somewhat visually appealing.

**Solution**

```{r,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
plot(eruptions ~ waiting, data = faithful,
xlab = "Waiting Time (in minutes)", ylab = "Eruption Duration (in minutes)", main = "Waiting Time vs Eruption Duration", pch = 20,
cex = 3,
col = "dodgerblue")
abline(faithful_model, lwd = 3, col = "darkorange")
```


**(g)** Report the value of $R^2$ for the model. Do so directly. Do not simply copy and paste the value from the full output in the console after running `summary()` in part **(a)**.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
SST = sum((faithful$eruptions - mean(faithful$eruptions)) ^ 2)
SSReg = sum((faithful_model$fitted.values - mean(faithful$eruptions)) ^ 2) 
R2 = SSReg/SST
R2
```


## Exercise 2 (Writing Functions)

This exercise is a continuation of Exercise 1.

**(a)** Write a function called `get_sd_est` that calculates an estimate of $\sigma$ in one of two ways depending on input to the function. The function should take two arguments as input:

- `model_resid` - A vector of residual values from a fitted model.
- `mle` - A logical (`TRUE` / `FALSE`) variable which defaults to `FALSE`.

The function should return a single value:

- $s_e$ if `mle` is set to `FALSE`.
- $\hat{\sigma}$ if `mle` is set to `TRUE`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
get_sd_est = function(model_resid, mle=FALSE){
SSE = sum(model_resid ^ 2)
if (mle == TRUE){
  sqrt(SSE / (length(model_resid) - 2))
} else {
  sqrt(SSE / length(model_resid))
}
}
```


**(b)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `FALSE`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
get_sd_est(faithful_model$residuals)
```

**(c)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `TRUE`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
get_sd_est(faithful_model$residuals,TRUE)
```

**(d)** To check your work, output `summary(faithful_model)$sigma`. It should match at least one of **(b)** or **(c)**.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
summary(faithful_model)$sigma
```

## Exercise 3 (Simulating SLR)

Consider the model

\[
Y_i = 3 - 7 x_i + \epsilon_i
\]

with 

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 4)
\]

where $\beta_0 = 3$ and $\beta_1 = -7$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous assignment.

```{r}
birthday = 19920120
set.seed(birthday)
```

**(a)** Use `R` to simulate `n = 50` observations from the above model. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 50, 0, 10)
```

You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Store the data frame this function returns in a variable of your choice. Note that this function calls $y$ `response` and $x$ `predictor`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sim_slr = function(n, x, beta_0, beta_1, sigma = 1) { epsilon = rnorm(n, mean = 0, sd = sigma)
y = beta_0 + beta_1 * x + epsilon 
data.frame(predictor = x, response = y)
}
x = runif(n = 50, 0, 10)
sim_data = sim_slr(n = 50, x = x, beta_0 = 3, beta_1 = -7, sigma = 2)
head(sim_data)
```




**(b)** Fit a model to your simulated data. Report the estimated coefficients. Are they close to what you would expect? Briefly explain.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
sim_fit = lm(response ~ predictor, data = sim_data)
coef(sim_fit)

```

They are pretty close to whats expected, negative slope and positive intercept of expected magnitudes. 

**(c)** Plot the data you simulated in part **(a)**. Add the regression line from part **(b)**. Hint: Keep the two commands in the same chunk, so `R` knows what plot to add the line to when knitting your `.Rmd` file.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
plot(response ~ predictor, data = sim_data, 
     xlab = "Simulated Predictor Variable", 
     ylab = "Simulated Response Variable", 
     main = "Simulated Regression Data", 
     pch = 20,cex = 2,col = "dodgerblue")
abline(sim_fit, lwd = 3, col = "darkorange")

```


**(d)** Use `R` to repeat the process of simulating `n = 50` observations from the above model $2000$ times. Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. Some hints:

- Use a `for` loop.
- Create `beta_hat_1` before writing the `for` loop. Make it a vector of length $2000$ where each element is `0`.
- Inside the body of the `for` loop, simulate new $y$ data each time. Use a variable to temporarily store this data together with the known $x$ data as a data frame.
- After simulating the data, use `lm()` to fit a regression. Use a variable to temporarily store this output.
- Use the `coef()` function and `[]` to extract the correct estimated coefficient.
- Use `beta_hat_1[i]` to store in elements of `beta_hat_1`.
- See the notes on [Distribution of a Sample Mean](http://daviddalpiaz.github.io/appliedstats/introduction-to-r.html#distribution-of-a-sample-mean) for some inspiration.

You can do this differently if you like. Use of these hints is not required.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
beta_hat_1 = apply(data.frame(rep(0,2000)), 1, function(y) coef(lm(response ~ predictor, data = sim_slr(n = 50, x = x, beta_0 = 3, beta_1 = -7, sigma = 2)))["predictor"])
head(beta_hat_1)
```


**(e)** Report the mean and standard deviation of `beta_hat_1`. Do either of these look familiar?

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
mean(beta_hat_1)
sd(beta_hat_1)
```
Mean looks familiar, approximately equal to $7$.

**(f)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

**Solution**
```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
hist(beta_hat_1, breaks = 50, 
      main = "Histogram of Estimated Regression Coeffecients",
      xlab = "Predictor Coeffecient")
```

The shape verifies CLT.

## Exercise 4 (Be a Skeptic)

Consider the model

\[
Y_i = 10 + 0 x_i + \epsilon_i
\]

with

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 1)
\]

where $\beta_0 = 10$ and $\beta_1 = 0$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous assignment.

```{r}
birthday = 19920120
set.seed(birthday)
```

**(a)** Use `R` to repeat the process of simulating `n = 25` observations from the above model $1500$ times. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 25, 0, 10)
```

Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Hint: Yes $\beta_1 = 0$.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
beta_hat_1 = apply(data.frame(rep(0,1500)), 1, function(y) coef(lm(response ~ predictor, data = sim_slr(n = 25, x = x, beta_0 = 10, beta_1 = 0, sigma = 1)))["predictor"])
head(beta_hat_1)
```

**(b)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
hist(beta_hat_1, breaks = 50, 
      main = "Histogram of Estimated Regression Coeffecients",
      xlab = "Predictor Coeffecient")
```

The shape verifies CLT.

**(c)** Import the data in [`skeptic.csv`](skeptic.csv) and fit a SLR model. The variable names in `skeptic.csv` follow the same convention as those returned by `sim_slr()`. Extract the fitted coefficient for $\beta_1$.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
skeptic <- read.csv("skeptic.csv")
beta_1_hat = coef(lm(response ~ predictor, data = skeptic))["predictor"]
beta_1_hat
```

**(d)** Re-plot the histogram from **(b)**. Now add a vertical red line at the value of $\hat{\beta_1}$ in part **(c)**. To do so, you'll need to use `abline(v = c, col = "red")` where `c` is your value.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
hist(beta_hat_1, breaks = 50, 
      main = "Histogram of Estimated Regression Coeffecients",
      xlab = "Predictor Coeffecient")
abline(v = beta_1_hat, col = "red")
```

**(e)** Your value of $\hat{\beta_1}$ in **(c)** should be positive. What proportion of the `beta_hat_1` values are larger than your $\hat{\beta_1}$? Return this proportion, as well as this proportion multiplied by `2`.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
prop = length(beta_hat_1[beta_hat_1 > beta_1_hat])/length(beta_hat_1)
prop
2*prop
```

**(f)** Based on your histogram and part **(e)**, do you think the [`skeptic.csv`](skeptic.csv) data could have been generated by the model given above? Briefly explain.

**Solution**
```{r,message=FALSE}
within_95 = beta_1_hat > (mean(beta_hat_1) + 2*sd(beta_hat_1))
within_95
```


The data in `skeptic.csv` could not have been generated by model given above, as the estimated coeffecient doesnt lie in $95\%$ CI.

## Exercise 5 (Comparing Models)

For this exercise we will use the data stored in [`goalies.csv`](goalies.csv). It contains career data for all 716 players in the history of the National Hockey League to play goaltender through the 2014-2015 season. The variables in the dataset are:

- `Player` - NHL Player Name
- `First` - First year of NHL career
- `Last` - Last year of NHL career
- `GP` - Games Played
- `GS` - Games Started
- `W` - Wins
- `L` - Losses
- `TOL` - Ties/Overtime/Shootout Losses
- `GA` - Goals Against
- `SA` - Shots Against
- `SV` - Saves
- `SV_PCT` - Save Percentage
- `GAA` - Goals Against Average
- `SO` - Shutouts
- `MIN` - Minutes
- `G` - Goals (that the player recorded, not opponents)
- `A` - Assists (that the player recorded, not opponents)
- `PTS` - Points (that the player recorded, not opponents)
- `PIM` - Penalties in Minutes

For this exercise we will define the "Root Mean Square Error" of a model as

\[
RMSE = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

**(a)** Fit a model with "wins"" as the response and "minutes" as the predictor. Calculate the RMSE of this model. Also provide a scatterplot with the fitted regression line.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
goalies <- read.csv("goalies.csv")
model = lm(W ~ MIN, data = goalies)
RMSE = sqrt(sum((model$residuals)^2)/length(goalies$W))
RMSE
plot(W ~ MIN, data = goalies,
xlab = "Minutes", ylab = "Wins", main = "Wins vs Minutes", pch = 20,
cex = 3,
col = "dodgerblue")
abline(model, lwd = 3, col = "darkorange")
```

**(b)** Fit a model with "wins"" as the response and "goals against" as the predictor. Calculate the RMSE of this model. Also provide a scatterplot with the fitted regression line.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
goalies <- read.csv("goalies.csv")
model = lm(W ~ GA, data = goalies)
RMSE = sqrt(sum((model$residuals)^2)/length(goalies$W))
RMSE
plot(W ~ GA, data = goalies,
xlab = "Goals Against", ylab = "Wins", main = "Wins vs Goals Against", pch = 20,
cex = 3,
col = "dodgerblue")
abline(model, lwd = 3, col = "darkorange")
```

**(c)** Fit a model with "wins"" as the response and "shutouts" as the predictor. Calculate the RMSE of this model. Also provide a scatterplot with the fitted regression line.

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE, warning=FALSE}
goalies <- read.csv("goalies.csv")
model = lm(W ~ SO, data = goalies)
RMSE = sqrt(sum((model$residuals)^2)/length(goalies$W))
RMSE
plot(W ~ SO, data = goalies,
xlab = "Shutouts", ylab = "Wins", main = "Wins vs Shutouts", pch = 20,
cex = 3,
col = "dodgerblue")
abline(model, lwd = 3, col = "darkorange")
```

**(d)** Based on the previous three models, which of the three predictors used is most helpful for predicting wins? Briefly explain.

**Solution**

Based on the previous three models, "minutes" is the best predictor for "wins" as it has least RMSE.

**(e)** *This question is not graded. You may delete it if you like.* How many games did Ed Belfour win? Why is he David Dalpiaz's favorite goalie on this list?

**Solution**

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
goalies[goalies$Player=="Ed Belfour*","W"]
```

He is ranked 3rd in number of Wins.
